{\bf \huge Wikiplia}:\\ The Free Programming Language\\ That Anyone Can Edit

### 2007-03-19 trunk/wikiplia/paper/wikiplia.tex
screenshot

\documentclass[twocolumn]{article}
\usepackage{fullpage}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{proof}
\usepackage{code}

@@

\newcommand\ftinker{\ensuremath{\copyright}}
\newcommand\fexpression{\ensuremath{\hbar}}
\newcommand\fbeer{\ensuremath{\triangle}}
\newcommand\fusa{\ensuremath{\sqrt[x^{1.2}]{\mbox{\t{oo}} + \frac{x}{z^2}}}}

@@

\paragraph{Freedom \fexpression: Freedom of expression.}
Programmers should be able to write their programs using any
expressions that they like. Specifically, there should be no
prior establishment of arbitrary categories of expression that
are excluded, such as those that discriminate on the basis
of class, XXXXX, or type.

@@

@@

It may na\"ively seem that publishing the entire CVS history of a
project would solve the issue with language extensions: By inspecting
the revision that introduced the {\tt compile\_a\_program} feature
(but prior to the replacement of the GCC with the minuscule version
above), one could then see its implementation and then know how to
HERE

% parallel development yields same source, different behavior.
% need globally atomic commits across all developers

All popular modern languages are defined via a definitional
interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
accompanying O'Reilly ``animal'' book.

@@

Revision control is accessed through the class of imperative {\tt cvs}
judgments. We assume a single global repository, which maps {\em keys}
(strings) to lists of {\em revisions}. A revision is a monotonically
increasing and unique {\em revision number} (integer) paired with an
XML document. The judgment ${\tt cvs commit}\ s\ X = i$ creates a new
revision with revision number $i$ and data $X$ and inserts it under
the key $s$. The judgment ${\tt cvs checkout}\ s = X$ fetches the most
recent revision for the key $s$, the document $X$ (if no such key
exists, then the document is invalid). The judgment ${\tt cvs\
checkout\ -\!r}\ i\ s = X$ does the same, but for the specific
revision number $i$ (if no such revision exists, the document is
invalid). Finally, ${\tt cvs\ log}\ s = \vec{X}$ fetches all of the
revision numbers for the key $s$, as a series of integers $\vec{X}$.

@@

\begin{figure}[b]
\begin{center}
\begin{tabular}{c}
  \begin{tabular}{rcl}
   {\em E} & ::= & {\tt(} $E_1\ E_2\ \ldots\ E_n$ {\tt)} \\
           & $|$ & {\tt"}string{\tt"} \\
           & $|$ & $n$ \\
           & $|$ & symbol \\
           & $|$ & {\tt'}E \\
  \end{tabular} \\[3em]

@@

XML documents are universally parseable.\z\wcite{Parsing} However,
they are difficult to write and read. Therefore, as usual\wcite{RELAX
NG} we create a new syntax by which humans can write and read
documents and which the computer automatically parses and converts to
the easily parsable XML syntax and back to the new syntax, reducing
complexity.\z\footnote{For efficiency, the Wikiplia implementation is
optimized to lazily perform these translations, so that the document
is never represented in XML form.} This compact syntax is based upon
parentheses rather than tags: The XML document $\tag{list}{X_1\ X_2}$
is instead written ${\tt(}X_1\ X_2{\tt)}$. Note that the closing
parenthesis is not {\em named} as in XML, which makes parsing
difficult because the computer must {\em guess} which parenthesis
belongs to which other parenthesis. We therefore call this compact
syntax XESP because it basically {\em reads the programmer's
mind}\wcite{Extra-sensory perception} to guess what the name of the
closing parenthesis should be. The grammar for XESP is given in
Figure~\ref{fig:xesp} along with the translation to XML documents.
From now on, we use the XESP syntax in our examples.

>>>>

\documentclass[twocolumn]{article}
\usepackage{fullpage}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{proof}
\usepackage{code}
\usepackage{graphicx}

@@

% TODO: license
\newcommand\currentrevision{465}

\newcommand\ftinker{\ensuremath{\copyright}}
\newcommand\fexpression{\ensuremath{\hbar}}
\newcommand\fbeer{\ensuremath{\triangle}}
\newcommand\fusa{\ensuremath{\sqrt[x^{1.2}]{\mbox{\t{oo}} + \frac{x}{z^2}}}}
\newcommand\ffree{\ensuremath{\P}}

@@

\paragraph{Freedom \fexpression: Freedom of expression.}
Programmers should be able to write their programs using any
expressions that they like. Specifically, there should be no
prior establishment of arbitrary categories of expression that
are excluded, such as those that discriminate on the basis
of class, mathematical philosophy, or type.

@@

\paragraph{Freedom \ffree: Freedom to redefine freedom.}
Freedom should be free, so the definition of freedom should be free to
change as the meaning of freedom changes. Wikiplia's license allows
for Wikiplia to be distributed in a way that monotonically increases
freedom as new concepts and catchphrases of freedom are invented.

@@

It may na\"\i{}vely seem that publishing the entire CVS history of a
project would solve the issue with language extensions: By inspecting
the revision that introduced the {\tt compile\_a\_program} feature
(but prior to the replacement of the GCC with the minuscule version
above), one could then see its implementation and then know what it
means. For certain patterns of development this does indeed suffice.
However, programmers are not forced to check in their changes except
at their own whims, as determined by social conventions; a programmer
might make the private addition of the keyword {\tt
compile\_a\_program}, then rewrite the GCC to use it, and only then
check in this change as one revision. For this action he will surely
be rebuked by his fellow programmers; none of the other developers can
compile the new version of the code without access to the intermediate
revision! This social pressure would also na\"\i{}vely seem to be enough
to address the problem, but more insidious scenarios yet obtain.

As a concrete example, suppose there are two programmers called K and
R. Each is modifying the GCC with the purpose of adding a new
character constant, \verb+'\c'+. K and R start at revision 100 of the
GCC. K finds the case analysis for parsing character constants:
%
\begin{code}
/* REVISION 100 (K) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  default: abort("bad char constant");
\}
\end{code}
%
He adds a case for his extension, without using the extension, and
checks this in as revision 101.
%
\begin{code}
/* REVISION 101 (K) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return 257;
  default: abort("bad char constant");
\}
\end{code}
%
Meanwhile, R has similar (but not identical) inspiration and modifies
his copy of the compiler:
%
\begin{code}
/* REVISION 100 (R) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return 8675309;
  default: abort("bad char constant");
\}
\end{code}
%
He does not commit his code because he is wary of the time-consuming
conflict resolution phase and is late for a date with K's estranged
wife who is fed up with K's all-night hacking binges. He takes off in
his 2007 Honda Civic\wcite{Honda Civic} with aftermarket spoiler for a
night on the town, believing that a healthy well-rounded programmer
spends more or less equal nights basking in the pale amber glow of the
teletype as waking up with a few missing teeth naked and norovirused
in some midtown alleyway with his wallet barely out of reach but empty
anyway, having amply exercised Freedom~\fbeer.

Meanwhile, K continues extending the GCC, using the extension to
implement itself. He checks in this code with no conflicts:
%
\begin{code}
/* REVISION 102 (K) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return '\\c';
  default: abort("bad char constant");
\}
\end{code}
%
K punches out at {\tt 1130 UTC},\z\wcite{Coordinated Universal Time}
just as R returns from his adventure. R's confidence bolstered, he
finishes his extension effort, following best practices and
implementing the extension using itself:
%
\begin{code}
/* REVISION 100 (R) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return '\\c';
  default: abort("bad char constant");
\}
\end{code}
%
He now decides to commit his changes (forgetting that he did not
commit the intermediate revision). To do so he updates to the newest
revision, 102, and sees that there are no conflicts---in fact,
revision 102 already contains his changes! Believing that his changes
are therefore compatible, he continues hacking.

After this scenario, K and R believe they are working on the same
programming language---after all, it has the same source code---but
their minor bifurcation in development history means that they forever
differing meanings of the \verb+'\c'+ extension. This mistake is
likely to go unnoticed for some time, and until it is resolved, the
meaning of the \verb+'\c'+ extension is firmly enslaved in the
bipartite penitentiary of double entendre, yearning to be
free\ldots\z\wcite{Information wants to be free}

\subsection{Solution}

Based on these scenarios we conclude that extant social measures such
as revision control conventions are not enough to guarantee freedom.
Even if we think these situations are implausible in the hands of
well-intentioned, well-mannered software engineers, we wish for our
software to remain free even when in the hands of nefarious and crafty
factions who would seek to fracture\wcite{Fork (software development)}
our free software community. We therefore need a technological and
legal solution that forces the entire development history to be
available. Keeping with Freedom~\fusa, we call this technology and its
license {\em Total Information Awareness} after the successful project
of the US Government with the same name.\z\wcite{Information Awareness
Office}

Technologically, we develop our system around a primitive notion of
revision control in which every change to the system is recorded.
Because the system has an integrated editor, every action of a
programmer is logged and preserved indefinitely with no extra action
necessary on the programmer's part. Such commits are globally atomic,
using a single universal repository. (This means that in the above
scenario, R would not have been able to forget to commit his
intermediate change, and would have been forced to observe his
conflict with K.) To protect against the possibility that divergent
development paths lead to incompatible compilers, we require that
there is only one compiler for any given programming language, which
itself exists in the revision control system. Therefore, it is always
clear which version of the compiler was used to produce an executable
from source. This also makes the K and R scenario impossible: there is
only one compiler and so it is impossible for it to differ from
itself.

Legally, all of the software is licensed (Section~\ref{sec:license})
under terms similar to the GNU GPL, but that define the ``source
code'' to include the entire revision history of the system. In order
to be compatible with Freedom~\ffree, we allow the license itself to be
edited, but to ensure that no one can remove freedoms already present
in the license, the license includes a provision that allows any {\em
prior} version of the license to be used, at the programmer's option.

\subsection{Implementation}

Wikiplia, the free programming language that anyone can edit, is
implemented as a web-site on the Internet at the address
$${\tt http://wikiplia.spacebar.org/}$$

\subsection{Roadmap}

The remainder of this paper proceeds as follows. We first discuss in
Section~\ref{sec:calculus} the design of the initial Wikiplia system,
which is used to bootstrap the rest of Wikiplia. We then discuss the
current state of Wikiplia as of revision \currentrevision\ in
Section~\ref{sec:current}. We discuss the freedom-preserving TIA
Public License in Section~\ref{sec:license}. We conclude with a
discussion of unrelated work and plans for the future~\ref{sec:conc}.

@@

Revision control is accessed through the class of imperative {\tt cvs}
judgments. We assume a single global repository, which maps {\em keys}
(strings) to lists of {\em revisions}. A revision is a monotonically
increasing and unique {\em revision number} (integer) paired with an
XML document. The judgment ${\tt cvs\ commit}\ s\ X = i$ creates a new
revision with revision number $i$ and data $X$ and inserts it under
the key $s$. The judgment ${\tt cvs\ checkout}\ s = X$ fetches the most
recent revision for the key $s$, the document $X$ (if no such key
exists, then the document is invalid). The judgment ${\tt cvs\ 
checkout\ -\!r}\ i\ s = X$ does the same, but for the specific
revision number $i$ (if no such revision exists, the document is
invalid). Finally, ${\tt cvs\ log}\ s = \vec{X}$ fetches all of the
revision numbers for the key $s$, as a series of integers $\vec{X}$.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
  \begin{tabular}{rcl}
   {\em E} & ::= & {\tt(} $E_1\ E_2\ \ldots\ E_n$ {\tt)} \\
           & $|$ & {\tt"}string{\tt"} \\
           & $|$ & $n$ \\
           & $|$ & symbol \\
           & $|$ & {\tt'}E \\
  \end{tabular} \\[3em]

@@

XML documents are universally parseable.\z\wcite{Parsing} However,
they are difficult to write and read. Therefore, as usual\wcite{RELAX
NG} we create a new syntax by which humans can write and read
documents and which the computer automatically parses and converts to
the easily parsable XML syntax and back to the new syntax, reducing
complexity.\z\footnote{For efficiency, the Wikiplia implementation is
optimized to lazily perform these translations, so that the document
is never represented in XML form.} This compact syntax is based upon
parentheses rather than tags: The XML document $\tag{list}{X_1\ X_2}$
is instead written ${\tt(}X_1\ X_2{\tt)}$. Note that the closing
parenthesis is not {\em named} as in XML, which makes parsing
difficult because the computer must {\em guess} which parenthesis
belongs to which other parenthesis. We therefore call this compact
syntax XESP because it basically {\em reads the programmer's
mind}\,\wcite{Extra-sensory perception} to guess what the name of the
closing parenthesis should be. The grammar for XESP is given in
Figure~\ref{fig:xesp} along with the translation to XML documents.
From now on, we use the XESP syntax in our examples.

\subsection{Implementation}
% SML: miserable non-free language

Wikiplia is implemented as a World Wide Web Home-Site, which allows
for easy access from any location.

\section{Revision \currentrevision} \label{sec:current}

\begin{figure*}[t]
\begin{center}
\includegraphics[width=0.92\textwidth]{ss_homepage}
\end{center}
\caption{Screenshot of the Wikiplia home page as of revision
\currentrevision.} \label{fig:sshomepage}
\end{figure*}

\section{TIA Public License} \label{sec:license}

In this section we reproduce the Total Information Awareness Public
License. Commentary is given via footnotes\wcite{footnote} into parts
of the license.

{\tt \parindent 0pt \parskip 10pt
\verb+     +TIA PUBLIC LICENSE \\
\verb+   +Revision 468, March 2007

BEGIN INVINCIBLE SECTION 1

This software is Copyright \copyright\ 2007{\large \rm--}$\infty$ The
Regents of the Wikiplia Foundation. Permission is not granted to
reproduce this software or license except by the terms explicitly
enumerated below.

I. Invincible Sections

This license contains certain invincible sections, denoted by the text
``BEGIN INVINCIBLE SECTION <n>'' and ``END INVINCIBLE SECTION <n>''.
Such sections may not be modified under any circumstances.

END INVINCIBLE SECTION 1\footnote{Invincible sections exist in order
to ensure the sanctity of the license. We first establish that
invincible sections will appear and that they are inviolable; this
itself is done in an invincible section. Invincible sections cause a
limited loss of liberty, but this is the cost of freedom.}

BEGIN INVINCIBLE SECTION 2

II. Version Idenfitication and Invalid Licenses

This license must identify itself in the header as Revision <n> for
some number <n>, which must be the same as the revision number in the
Wikiplia repository for the key ``TPL'' in which the license text is
stored. If this is not the case, then this version of the license is
considered Invalid and Void.

Permission is not granted to distribute the software or license using
any Invalid version of the license.

END INVINCIBLE SECTION 2\footnote{This invincible section establishes
a connection between license versions and the actual contents of the
Wikiplia repository. Note the self-reference: Though this text is in
an invincible section and never changes, the referent of ``this
license'' does change as the rest of the license is modified. Because
Wikiplia assigns version numbers monotonically, this ensures that the
next invincible section is able to guarantee that freedom is
monotonic. In the case of an invalid license, no permissions
whatsoever are granted, so the licensee {\em must} use a prior valid
version of the license. The initial version of the license is valid.}

BEGIN INVINCIBLE SECTION 3

III. Option of License

The licensee has the option to choose any revision of the license
prior to (numerically less than) this version as the licensing
terms for the software and license.

END INVINCIBLE SECTION 3

BEGIN INVINCIBLE SECTION 4

IV. Heredity of License

Any copy or derivative work of this software or license must be
licensed under the TIA Public License.

END INVINCIBLE SECTION 4\footnote{This clause makes the license
``viral'' like the GNU GPL, so that freedom is preserved in all
descendents of the software.}

BEGIN INVINCIBLE SECTION 5

V. Completeness of Copy

This software and license may only be copied in their entirety,
including the entire revision history.

END INVINCIBLE SECTION 5\footnote{This section is the centerpiece of
Wikiplia; it guarantees that the ``source code'' to any software or
programming language derived from Wikiplia is free from the loopholes
described in Section~\ref{sec:bootstrap} and so maximizes
Freedom~\ftinker. Note that this does not limit the way that the
software can be modified; the programmer might begin by blanking all
of the keys he doesn't care about---as long as he preserves the fact
that there was {\em once} something there.}

VI. Freedom to edit

Permission is hereby granted to edit this license.\footnote{The only
non-invincible provision of the original license allows the reader to
add provisions that he desires to the license. This guarantees
Freedom~\ffree, the freedom to redefine freedom.
%
Note that even if a freedom-hater removes this provision from the
license, the invincible sections above ensure that the freedom to edit
the license is preserved for all time.}

}

\bigskip
This license is bootstrapping in the sense that it grants only the
minimal permissions necessary, after setting up invariants via the
behavior-limiting invincible sections. In fact, the original version
of the license does not directly permit the licensee to copy the
software at all; he must first amend the license using provision {\tt
VI} to give himself this permission.

\section{Conclusion} \label{sec:conc}

% All popular modern languages are defined via a definitional
% interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
% accompanying O'Reilly ``animal'' book.
